Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-18 talk/pepm2011/paper.tex
comments by stephan in section 5

\begin{figure*}
\begin{center}
\begin{tabular}{lcccc}
\emph{new} & ${\displaystyle \frac{l\,\mathrm{fresh}}{v=\mathtt{new}(T),E,H\overset{\mathrm{run}}{\Longrightarrow}E\left[v\mapsto l\right],H\left[l\mapsto\left(T,\mathrm{null},\mathrm{null}\right)\right]}}$ & ~~~ & \emph{guard} & ${\displaystyle \frac{\mathrm{type}(H(E(v))=T}{\mathtt{guard\_class}(v,T),E,H\overset{\mathrm{run}}{\Longrightarrow}E,H}}$\tabularnewline[3em]
\emph{get} & ${\displaystyle \frac{\,}{u=\mathtt{get}(v,F),E,H\overset{\mathrm{run}}{\Longrightarrow}E\left[u\mapsto H\left(E\left(v\right)\right)_{F}\right],H}}$ & ~~~ &  & ${\displaystyle \frac{\mathrm{type}(H(E(v))\neq T}{\mathtt{guard\_class}(v,T),E,H\overset{\mathrm{run}}{\Longrightarrow}\bot,\bot}}$\tabularnewline[3em]
\emph{set} & ${\displaystyle \frac{\,}{\mathtt{set}\left(v,F,u\right),E,H\overset{\mathrm{run}}{\Longrightarrow}E,H\left[E\left(v\right)\mapsto\left(H\left(E\left(v\right)\right)!_{F}E(u)\right)\right]}}$ & ~~~ &  & \tabularnewline[4em]
\end{tabular}

\begin{minipage}[b]{7 cm}
\emph{Object Domains:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    u,v & \in & V & \mathrm{\ variables\ in\ trace}\\
    T & \in & \mathfrak{T} & \mathrm{\ runtime\ types}\\
    F & \in & \left\{ L,R\right\} & \mathrm{\ fields\ of\ objects}\\
    l & \in & L & \mathrm{\ locations\ on\ heap}
 \end{array}
$$
\end{minipage}
\begin{minipage}[b]{5 cm}
\emph{Semantic Values:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    E & \in & V\rightharpoonup L & \mathrm{\ Environment}\\
    H & \in & L\rightharpoonup\mathfrak{T}\times(L\cup \{ \mathrm{null} \})\times (L\cup \{ \mathrm{null}\} ) & \mathrm{\ Heap}\\
    \\
    \\
 \end{array}
$$
\end{minipage}
\end{center}
\caption{The Operational Semantics of Simplified Traces}
\label{fig:semantics}
\end{figure*}

In this section we want to give a formal description of the semantics of the
traces and of the optimizer and liken the optimization to partial evaluation.
We concentrate on the operations for manipulating dynamically allocated objects,
as those are the only ones that are actually optimized. Without loss of
generality we also consider only objects with two fields in this section.

Traces are lists of operations. The operations considered here are \lstinline{new} (to make
a new object), \lstinline{get} (to read a field out of an object), \lstinline{set} (to write a field
into an object) and \lstinline{guard_class} (to check the type of an object). The values of all
variables are locations (i.e.~pointers). Locations are mapped to objects, which
are represented by triples of a type $T$, and two locations that represent the
fields of the object. When a new object is created, the fields are initialized
to null, but we require that they are initialized to a real
location before being read, otherwise the trace is malformed.

@@

Figure~\ref{fig:semantics} shows the operational semantics for traces. The
interpreter formalized there executes one operation at a time. Its state is
represented by an environment $E$ and a heap $H$, which are potentially changed by the
execution of an operation. The environment is a partial function from variables
to locations and the heap is a partial function from locations to objects. Note
that a variable can never be null in the environment, otherwise the trace would
be malformed. The environment could not directly map variables to object,
because several variables can contain a pointer to the \emph{same} object. 
The "indirection" is needed to express sharing.

@@

\begin{figure*}
\begin{center}
\begin{tabular}{lc}
\emph{new} & ${\displaystyle \frac{v^{*}\,\mathrm{fresh}}{v=\mathtt{new}(T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E\left[v\mapsto v^{*}\right],S\left[v^{*}\mapsto\left(T,\mathrm{null,null}\right)\right]}}$\tabularnewline[3em]
\emph{get} & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S)}{u=\mathtt{get}(v,F),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E\left[u\mapsto S(E(v))_{F}\right],S}}$\tabularnewline[3em]
 & ${\displaystyle \frac{E(v)\notin\mathrm{dom}(S)\, u^{*}\,\mathrm{fresh}}{u=\mathtt{get}(v,F),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle u^{*}=\mathtt{get}(E(v),F)\right\rangle ,E\left[u\mapsto u^{*}\right],S}}$\tabularnewline[3em]
\emph{set} & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S)}{\mathtt{set}\left(v,F,u\right),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E,S\left[E\left(v\right)\mapsto\left(S(E(v))!_{F}E(u)\right)\right]}}$\tabularnewline[3em]
 & ${\displaystyle \frac{E(v)\notin\mathrm{dom}\left(S\right),\,\left(E(v),S\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{\mathtt{set}\left(v,F,u\right),E,S\overset{\mathrm{opt}}{\Longrightarrow}\mathrm{ops}::\left\langle \mathtt{set}\left(E(v),F,E(u)\right)\right\rangle ,E,S^{\prime}}}$\tabularnewline[3em]
\emph{guard} & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S),\,\mathrm{type}(S(E(v)))=T}{\mathtt{guard\_class}(v,T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E,S}}$\tabularnewline[3em]
 & ${\displaystyle \frac{E(v)\notin\mathrm{dom}(S)\vee\mathrm{type}(S(E(v)))\neq T,\,\left(E(v),S\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{\mathtt{guard\_class}(v,T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \mathtt{guard\_class}(E\left(v\right),T)\right\rangle ,E,S^{\prime}}}$\tabularnewline[3em]
\emph{lifting} & ${\displaystyle \frac{v^{*}\notin\mathrm{dom}(S)}{v^{*},S\overset{\mathrm{lift}}{\Longrightarrow}\left\langle \,\right\rangle ,S}}$\tabularnewline[3em]
 & ${\displaystyle \frac{v^{*}\in\mathrm{dom}(S),\,\left(v^{*},S\right)\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{v^{*},S\overset{\mathrm{lift}}{\Longrightarrow}\left\langle v^{*}=\mathtt{new}\left(T\right)\right\rangle ::ops,S^{\prime}}}$\tabularnewline[3em]
 & ${\displaystyle \frac{\left(S\left(v^{*}\right)_{L},S\setminus\left\{ v^{*}\mapsto S\left(v^{*}\right)\right\} \right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{L},S^{\prime}\right),\,\left(S\left(v^{*}\right)_{R},S^{\prime}\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{R},S^{\prime\prime}\right)}{v^{*},S\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\mathrm{ops}_{L}::ops_{R}::\left\langle \mathtt{set}\left(v^{*},L,S\left(v^{*}\right)_{L}\right),\,\mathtt{set}\left(v^{*},R,S\left(v^{*}\right)_{R}\right)\right\rangle ,S^{\prime}}}$\tabularnewline[3em]
\end{tabular}

\begin{minipage}[b]{7 cm}
\emph{Object Domains:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    u,v & \in & V & \mathrm{\ variables\ in\ trace}\\
    u^*,v^* & \in & V^* & \mathrm{\ variables\ in\ optimized\ trace}\\
    T & \in & \mathfrak{T} & \mathrm{\ runtime\ types}\\
    F & \in & \left\{ L,R\right\} & \mathrm{\ fields\ of\ objects}\\
 \end{array}
$$
\end{minipage}
\begin{minipage}[b]{5 cm}
\emph{Semantic Values:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    E & \in & V\rightharpoonup V^* & \mathrm{\ Environment}\\
    S & \in & V^*\rightharpoonup\mathfrak{T}\times(V^*\cup \{ \mathrm{null} \})\times (V^*\cup \{ \mathrm{null}\} ) & \mathrm{\ Static\ Heap}\\
    \\
    \\
 \end{array}
$$
\end{minipage}
\end{center}
\caption{Optimization Rules}
\label{fig:optimization}
\end{figure*}

To optimize the simple traces from the last section, we use online partial
evaluation. The partial evaluator optimizes one operation of the trace at a
time. Every operation in the unoptimized trace is replaced by a list of
operations in the optimized trace. This list is empty if the operation
can be optimized away (which hopefully happens often). The optimization rules
can be seen in Figure~\ref{fig:optimization}.

The state of the optimizer is stored in an environment $E$ and a \emph{static
heap} $S$. The environment is a partial function from variables in the
unoptimized trace $V$ to variables in the optimized trace $V^*$ (which are
themselves written with a
$\ ^*$ for clarity). The reason for introducing new variables in the optimized
trace is that several variables that appear in the unoptimized trace can turn
into the same variables in the optimized trace. The environment of the
optimizer serves a function similar to that of the environment in the semantics: sharing.

The static heap is a partial function from $V^*$ into the
set of static objects, which are triples of a type and two elements of $V^*$.
A variable $v^*$ is in the domain of the static heap $S$ as long as the
optimizer can fully keep track of the object. The object $S(v^*)$ is what is
statically known about the object stored in it, \ie its type and its fields. The
fields of objects in the static heap are also elements of $V^*$ (or null, for
short periods of time).

When the optimizer sees a \lstinline{new} operation, it optimistically removes it and
assumes that the resulting object can stay static. The optimization for all
further operations is split into two cases. One case is for when the
involved variables are in the static heap, which means that the operation can be
performed at optimization time and removed from the trace. These rules mirror
the execution semantics closely. The other case is for when not enough is known about
the variables, and the operation has to be residualized.

@@

If the first argument $v$ to a \lstinline{set} operation is mapped to something in the
static heap, then the \lstinline{set} can performed at optimization time and the static heap
updated. Otherwise the \lstinline{set} operation needs to be residualized. This needs to be
done carefully, because the new value for the field, from the variable $u$,
could itself be static, in which case it needs to be lifted first.

@@

Lifting takes a variable that is potentially in the static heap and makes sure
that it is turned into a dynamic variable. This means that operations are
emitted that construct an object with the shape described in the
static heap, and the variable is removed from the static heap.

@@

As an example for lifting, consider the static heap $$\{v^* \mapsto (T_1, w^*,
v^*), w^* \mapsto (T_2, u^*, u^*)\}$$ which contains two static objects. If $v^*$
now needs to be lifted, the following residual operations are produced:

@@

While we do not offer a formal proof of it, it should be relatively clear
that the algorithm presented above is sound: it works by delaying (and
often completely removing) some operations.  The algorithm runs in a
single pass over the list of operations.  We can check that although
recursively lifting a static object is not a constant-time operation,
the algorithm only takes a total time linear in the length of the trace.
The algorithm itself is not particularly complex; our focus is
rather that \emph{in the context of tracing JITs} it is possible to find a
simple enough algorithm that still gives very good results.

Note in particular that objects in category 1 (\ie the ones that do
not escape) are completely removed; moreover, objects in category 2
(\ie escaping) are still partially dealt with: if such an object
escapes later than its creation point, all the operations in between that
involve the object are removed.

The optimization is particularly effective for chains of operations.
For example, it is typical for an interpreter to generate sequences of
writes-followed-by-reads, where one interpreted opcode writes to some
object's field and the next interpreted opcode reads it back, possibly
dispatching on the type of the object created just before.  A typical example
would be chains of arithmetic operations.

@@

The allocation removal techniques described in this paper were implemented in
the optimizer of PyPy's tracing JIT. The optimization is independent of which
interpreter a JIT is generated for. There are some practical issues beyond the
techniques described in this paper. The actual implementation needs to deal with
more operations than described in Section~\ref{sec:formal}, for example to
also support static arrays in addition to static objects. The implementation of
this optimization is about 400 lines of RPython code.

A further complication is that most interpreters written with PyPy use
heap-allocated frame objects to store local variables. Those severely hinder
the effectiveness of allocation removal, because every time an object is stored
into a local variable, it is stored into the frame-object, which makes it
escape. We implemented a technique to treat such frames objects in a
special way to solve this problem. This is a common approach in VM
implementations \cite{miranda_context_1999,andreas_gal_trace-based_2009}; the
novelty of our approach is that we generalized it enough to be usable for
different interpreter.

@@

As the first step, we counted the occurring operations in all generated traces
before and after the optimization phase for all benchmarks. The resulting
numbers can be
seen in Figure~\ref{fig:numops}. The optimization removes as many as 90\% and as
little as 4\% percent of allocation operations in the traces of the benchmarks.
All benchmarks taken together, the optimization removes 70\% percent of
allocation operations. The numbers look similar for reading and writing of
attributes. There are even more \lstinline{guard} operations that are removed,
however there is an additional optimization that removes guards, so not all the
removed guards are an effect of the optimization described here.

\begin{figure*}
\begin{center}
\begin{tabular}{|l||r|rr|rr|rr|rr|}
\hline
	&num loops      &new    &removed        &get/set        &removed &guard &removed        &all ops        &removed\\
\hline
crypto\_pyaes	&78	&3088	&50\%	&57148	&25\%	&9055	&95\%	&137189	&80\%\\
django	&51	&673	&54\%	&19318	&18\%	&3876	&93\%	&55682	&85\%\\
fannkuch	&43	&171	&49\%	&886	&63\%	&1159	&81\%	&4935	&45\%\\
go	&517	&12234	&76\%	&200842	&21\%	&53138	&90\%	&568542	&84\%\\
html5lib	&498	&14432	&68\%	&503390	&11\%	&71592	&94\%	&1405780	&91\%\\
meteor-contest	&59	&277	&36\%	&4402	&31\%	&1078	&83\%	&12862	&68\%\\
nbody	&13	&96	&38\%	&443	&69\%	&449	&78\%	&2107	&38\%\\
pyflate-fast	&162	&2278	&55\%	&39126	&20\%	&8194	&92\%	&112857	&80\%\\
raytrace-simple	&120	&3118	&59\%	&91982	&15\%	&13572	&95\%	&247436	&89\%\\
richards	&87	&844	&4\%	&49875	&22\%	&4130	&91\%	&133898	&83\%\\
spambayes	&314	&5608	&79\%	&117002	&11\%	&25313	&94\%	&324125	&90\%\\
spectral-norm	&38	&360	&64\%	&5553	&20\%	&1122	&92\%	&11878	&77\%\\
telco	&46	&1257	&90\%	&37470	&3\%	&6644	&99\%	&98590	&97\%\\
twisted-names	&214	&5273	&84\%	&100010	&10\%	&23247	&96\%	&279667	&92\%\\
\hline
total	&2240	&49709	&70\%	&1227447	&14\%	&222569	&93\%	&3395548	&89\%\\
\hline
\end{tabular}
\end{center}
\caption{Number of Operations and Percentage Removed By Optimization}
\label{fig:numops}
\end{figure*}

@@

\section{Conclusion}
\label{sec:conclusion}

In this paper, we used an approach based on online partial evaluation
to optimize away allocations and type guards in the traces of a
tracing JIT.  In this context a simple approach to partial evaluation
gives good results.  This is due to the fact that the tracing JIT
itself is responsible for all control issues, which are usually the
hardest part of partial evaluation: the tracing JIT selects the parts
of the program that are worthwhile to optimize, and extracts linear
paths through them, inlining functions as necessary.  What is left to
optimize are only those linear paths.

@@

The authors would like to thank Stefan Hallerstede, David Schneider and Thomas
Stiehl for fruitful discussions during the writing of the paper.

>>>>

\begin{figure*}
\begin{center}
\begin{tabular}{lcccc}
\emph{new} & ${\displaystyle \frac{l\,\mathrm{fresh}}{v=\mathtt{new}(T),E,H\overset{\mathrm{run}}{\Longrightarrow}E\left[v\mapsto l\right],H\left[l\mapsto\left(T,\mathrm{null},\mathrm{null}\right)\right]}}$ & ~~~ & \emph{guard} & ${\displaystyle \frac{\mathrm{type}(H(E(v))=T}{\mathtt{guard\_class}(v,T),E,H\overset{\mathrm{run}}{\Longrightarrow}E,H}}$\tabularnewline[3em]
\emph{get} & ${\displaystyle \frac{\,}{u=\mathtt{get}(v,F),E,H\overset{\mathrm{run}}{\Longrightarrow}E\left[u\mapsto H\left(E\left(v\right)\right)_{F}\right],H}}$ & ~~~ &  & ${\displaystyle \frac{\mathrm{type}(H(E(v))\neq T}{\mathtt{guard\_class}(v,T),E,H\overset{\mathrm{run}}{\Longrightarrow}\bot,\bot}}$\tabularnewline[3em]
\emph{set} & ${\displaystyle \frac{\,}{\mathtt{set}\left(v,F,u\right),E,H\overset{\mathrm{run}}{\Longrightarrow}E,H\left[E\left(v\right)\mapsto\left(H\left(E\left(v\right)\right)!_{F}E(u)\right)\right]}}$ & ~~~ &  & \tabularnewline[4em]
\end{tabular}
\end{center}

\begin{minipage}[b]{7 cm}
\emph{Object Domains:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    u,v & \in & V & \mathrm{\ variables\ in\ trace}\\
    T & \in & \mathfrak{T} & \mathrm{\ runtime\ types}\\
    F & \in & \left\{ L,R\right\} & \mathrm{\ fields\ of\ objects}\\
    l & \in & L & \mathrm{\ locations\ on\ heap}
 \end{array}
$$
\end{minipage}
\begin{minipage}[b]{5 cm}
\emph{Semantic Values:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    E & \in & V\rightharpoonup L & \mathrm{\ Environment}\\
    H & \in & L\rightharpoonup\mathfrak{T}\times(L\cup \{ \mathrm{null} \})\times (L\cup \{ \mathrm{null}\} ) & \mathrm{\ Heap}\\
    \\
    \\
 \end{array}
$$
\end{minipage}
\caption{The Operational Semantics of Simplified Traces}
\label{fig:semantics}
\end{figure*}

In this section we want to give a formal description of the semantics of the
traces and of the optimizer and liken the optimization to partial evaluation.
We focus on the operations for manipulating dynamically allocated objects,
as those are the only ones that are actually optimized. Without loss of
generality we also consider only objects with two fields in this section.

Traces are lists of operations. The operations considered here are
\lstinline{new}, \lstinline{get}, \lstinline{set} and \lstinline{guard_class}.
The values of all
variables are locations (\ie pointers). Locations are mapped to objects, which
are represented by triples of a type $T$, and two locations that represent the
fields of the object. When a new object is created, the fields are initialized
to null, but we require that they are initialized to a real
location before being read, otherwise the trace is malformed (this condition is
guaranteed by how the traces are generated in PyPy).

@@

Figure~\ref{fig:semantics} shows the operational semantics for traces. The
interpreter formalized there executes one operation at a time. Its state is
represented by an environment $E$ and a heap $H$, which may be changed by the
execution of an operation. The environment is a partial function from variables
to locations and the heap is a partial function from locations to objects. Note
that a variable can never be null in the environment, otherwise the trace would have
been malformed. The environment could not directly map variables to objects,
because several variables can point to the \emph{same} object, because of aliasing.

@@

\begin{figure*}
\begin{center}
\begin{tabular}{lc}
\emph{new} & ${\displaystyle \frac{v^{*}\,\mathrm{fresh}}{v=\mathtt{new}(T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E\left[v\mapsto v^{*}\right],S\left[v^{*}\mapsto\left(T,\mathrm{null,null}\right)\right]}}$\tabularnewline[3em]
\emph{get} & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S)}{u=\mathtt{get}(v,F),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E\left[u\mapsto S(E(v))_{F}\right],S}}$\tabularnewline[3em]
 & ${\displaystyle \frac{E(v)\notin\mathrm{dom}(S),\,\,\, u^{*}\,\mathrm{fresh}}{u=\mathtt{get}(v,F),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle u^{*}=\mathtt{get}(E(v),F)\right\rangle ,E\left[u\mapsto u^{*}\right],S}}$\tabularnewline[3em]
\emph{set} & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S)}{\mathtt{set}\left(v,F,u\right),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E,S\left[E\left(v\right)\mapsto\left(S(E(v))!_{F}E(u)\right)\right]}}$\tabularnewline[3em]
 & ${\displaystyle \frac{E(v)\notin\mathrm{dom}\left(S\right),\,\left(E(v),S\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{\mathtt{set}\left(v,F,u\right),E,S\overset{\mathrm{opt}}{\Longrightarrow}\mathrm{ops}::\left\langle \mathtt{set}\left(E(v),F,E(u)\right)\right\rangle ,E,S^{\prime}}}$\tabularnewline[3em]
\emph{guard} & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S),\,\mathrm{type}(S(E(v)))=T}{\mathtt{guard\_class}(v,T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E,S}}$\tabularnewline[3em]
 & ${\displaystyle \frac{E(v)\notin\mathrm{dom}(S)\vee\mathrm{type}(S(E(v)))\neq T,\,\left(E(v),S\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{\mathtt{guard\_class}(v,T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\mathrm{ops}::\left\langle \mathtt{guard\_class}(E\left(v\right),T)\right\rangle ,E,S^{\prime}}}$\tabularnewline[3em]
\emph{lifting} & ${\displaystyle \frac{v^{*}\notin\mathrm{dom}(S)}{v^{*},S\overset{\mathrm{lift}}{\Longrightarrow}\left\langle \,\right\rangle ,S}}$\tabularnewline[3em]
 & ${\displaystyle \frac{v^{*}\in\mathrm{dom}(S),\,\left(v^{*},S\right)\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{v^{*},S\overset{\mathrm{lift}}{\Longrightarrow}\left\langle v^{*}=\mathtt{new}\left(T\right)\right\rangle ::ops,S^{\prime}}}$\tabularnewline[3em]
 & ${\displaystyle \frac{\left(S\left(v^{*}\right)_{L},S\setminus\left\{ v^{*}\mapsto S\left(v^{*}\right)\right\} \right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{L},S^{\prime}\right),\,\left(S\left(v^{*}\right)_{R},S^{\prime}\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{R},S^{\prime\prime}\right)}{v^{*},S\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\mathrm{ops}_{L}::ops_{R}::\left\langle \mathtt{set}\left(v^{*},L,S\left(v^{*}\right)_{L}\right),\,\mathtt{set}\left(v^{*},R,S\left(v^{*}\right)_{R}\right)\right\rangle ,S^{\prime}}}$\tabularnewline[3em]
\end{tabular}
\end{center}

\begin{minipage}[b]{7 cm}
\emph{Object Domains:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    u,v & \in & V & \mathrm{\ variables\ in\ trace}\\
    u^*,v^* & \in & V^* & \mathrm{\ variables\ in\ optimized\ trace}\\
    T & \in & \mathfrak{T} & \mathrm{\ runtime\ types}\\
    F & \in & \left\{ L,R\right\} & \mathrm{\ fields\ of\ objects}\\
 \end{array}
$$
\end{minipage}
\begin{minipage}[b]{5 cm}
\emph{Semantic Values:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    E & \in & V\rightharpoonup V^* & \mathrm{\ Environment}\\
    S & \in & V^*\rightharpoonup\mathfrak{T}\times(V^*\cup \{ \mathrm{null} \})\times (V^*\cup \{ \mathrm{null}\} ) & \mathrm{\ Static\ Heap}\\
    \\
    \\
 \end{array}
$$
\end{minipage}
\caption{Optimization Rules}
\label{fig:optimization}
\end{figure*}

To optimize the simple traces of the last section, we use online partial
evaluation. The partial evaluator optimizes one operation of a trace at a
time. Every operation in the unoptimized trace is replaced by a list of
operations in the optimized trace. This list is empty if the operation
can be optimized away. The optimization rules can be seen in
Figure~\ref{fig:optimization}.

The state of the optimizer is stored in an environment $E$ and a \emph{static
heap} $S$. The environment is a partial function from variables in the
unoptimized trace $V$ to variables in the optimized trace $V^*$ (which are
themselves written with a
$\ ^*$ for clarity). The reason for introducing new variables in the optimized
trace is that several variables that appear in the unoptimized trace can turn
into the same variables in the optimized trace. The environment of the
optimizer serves a function similar to that of the environment in the
semantics: to express sharing.

The static heap is a partial function from $V^*$ into the
set of static objects, which are triples of a type and two elements of $V^*$.
A variable $v^*$ is in the domain of the static heap $S$ as long as the
optimizer does not need to become dynamic XXX. The object $S(v^*)$ describes
what is statically known about the object, \ie its type and its fields. The
fields of objects in the static heap are also elements of $V^*$ (or null, for
short periods of time).

When the optimizer sees a \lstinline{new} operation, it optimistically removes it and
assumes that the resulting object can stay static. The optimization for all
further operations is split into two cases. One case is for when the
involved variables are in the static heap, which means that the operation can be
performed at optimization time and can be removed from the trace. These rules mirror
the execution semantics closely. The other case is for when not enough is known about
the variables, and the operation has to be residualized.

@@

If the first argument $v$ to a \lstinline{set} operation is mapped to something in the
static heap, then the \lstinline{set} can be performed at optimization time
(which updates the static heap). Otherwise the \lstinline{set} operation needs
to be residualized. This needs to be
done carefully, because the new value for the field, from the variable $u$,
could itself be static, in which case it needs to be lifted first.

@@

Lifting takes a variable and turns it into a dynamic variable. If the variable
is already dynamic, nothing needs to be done. If it is in the static heap,
operations are emitted that construct an object with the shape described
there, and the variable is removed from the static heap.

@@

As an example for lifting, consider the static heap $$\{v^* \mapsto (T_1, w^*,
v^*), w^* \mapsto (T_2, u^*, u^*)\}$$ which contains two static objects. If $v^*$
needs to be lifted, the following residual operations are produced:

@@

While we do not offer a formal proof of it, it should be relatively clear
that the algorithm presented above is sound: it works by delaying (and
often completely removing) some operations.  The algorithm runs in a
single pass over the list of operations.  We can check that although
recursively lifting a static object is not a constant-time operation,
the algorithm only takes a total time linear in the length of the trace.
The algorithm itself is not particularly complex; our focus is
rather that \emph{in the context of tracing JITs} it is possible to find a
simple enough algorithm that performs well.

Note in particular that objects in category 1 (\ie those that do
not escape) are completely removed; moreover, objects in category 2
(\ie escaping) are still partially optimized: all the operations in between the
creation of the object and the point where it escapes that involve the object
are removed.

The optimization is particularly effective for chains of operations.
For example, it is typical for an interpreter to generate sequences of
writes-followed-by-reads, where one interpreted opcode writes to some
object's field and the next interpreted opcode reads it back, possibly
dispatching on the type of the object created just before.  A typical example
would be a chain of arithmetic operations.

@@

The allocation removal techniques described in this paper were implemented in
the optimizer of PyPy's tracing JIT. The optimization is independent of which
interpreter a JIT is generated for. There are some practical issues beyond the
techniques described in this paper. The actual implementation needs to deal with
more operations than described in Section~\ref{sec:formal}, \eg to
also support static arrays in addition to static objects. The implementation of
this optimization is about 400 lines of RPython code.

A further complication is that most interpreters written with PyPy use
heap-allocated frame objects to store local variables. Those severely hinder
the effectiveness of allocation removal, because every time an object is stored
into a local variable, it is stored into the frame-object, which makes it
escape. We implemented a technique to treat such frames objects in a
special way to solve this problem. This is a common approach in VM
implementations \cite{miranda_context_1999,andreas_gal_trace-based_2009}; the
novelty of our approach is that we generalized it enough to be usable for
different interpreters.

@@

As the first step, we counted the occurring operations in all generated traces
before and after the optimization phase for all benchmarks. The resulting
numbers can be
seen in Figure~\ref{fig:numops}. The optimization removes between 4\% and 90\%
and of allocation operations in the traces of the benchmarks.
All benchmarks taken together, the optimization removes 70\% percent of
allocation operations. The numbers look similar for reading and writing of
attributes. There are even more \lstinline{guard} operations that are removed,
however there is an additional optimization that removes guards, so not all the
removed guards are an effect of the optimization described here (for technical
reasons, it would be very hard to separate the two effects).

\begin{figure*}
{\small
\begin{center}
\begin{tabular}{|l||r|rr|rr|rr|rr|}
\hline
	&num loops      &new    &removed        &get/set        &removed &guard &removed        &all ops        &removed\\
\hline
crypto\_pyaes	&78	&3088	&50\%	&57148	&25\%	&9055	&95\%	&137189	&80\%\\
django	&51	&673	&54\%	&19318	&18\%	&3876	&93\%	&55682	&85\%\\
fannkuch	&43	&171	&49\%	&886	&63\%	&1159	&81\%	&4935	&45\%\\
go	&517	&12234	&76\%	&200842	&21\%	&53138	&90\%	&568542	&84\%\\
html5lib	&498	&14432	&68\%	&503390	&11\%	&71592	&94\%	&1405780	&91\%\\
meteor-contest	&59	&277	&36\%	&4402	&31\%	&1078	&83\%	&12862	&68\%\\
nbody	&13	&96	&38\%	&443	&69\%	&449	&78\%	&2107	&38\%\\
pyflate-fast	&162	&2278	&55\%	&39126	&20\%	&8194	&92\%	&112857	&80\%\\
raytrace-simple	&120	&3118	&59\%	&91982	&15\%	&13572	&95\%	&247436	&89\%\\
richards	&87	&844	&4\%	&49875	&22\%	&4130	&91\%	&133898	&83\%\\
spambayes	&314	&5608	&79\%	&117002	&11\%	&25313	&94\%	&324125	&90\%\\
spectral-norm	&38	&360	&64\%	&5553	&20\%	&1122	&92\%	&11878	&77\%\\
telco	&46	&1257	&90\%	&37470	&3\%	&6644	&99\%	&98590	&97\%\\
twisted-names	&214	&5273	&84\%	&100010	&10\%	&23247	&96\%	&279667	&92\%\\
\hline
total	&2240	&49709	&70\%	&1227447	&14\%	&222569	&93\%	&3395548	&89\%\\
\hline
\end{tabular}
\end{center}
}
\caption{Number of Operations and Percentage Removed By Optimization}
\label{fig:numops}
\end{figure*}

@@

\section{Conclusion and Future Work}
\label{sec:conclusion}

In this paper, we used an approach based on online partial evaluation
to optimize away allocations and type guards in the traces of a
tracing JIT.  In this context a simple approach based on partial evaluation
gives good results.  This is due to the fact that the tracing JIT
itself is responsible for all control issues, which are usually the
hardest part of partial evaluation: the tracing JIT selects the parts
of the program that are worthwhile to optimize, and extracts linear
paths through them, inlining functions as necessary.  What is left to
optimize are only those linear paths.

@@

The authors would like to thank Stefan Hallerstede, David Schneider and Thomas
Stiehl for fruitful discussions and detailed feedback during the writing of the
paper.
